---
title: "Lecture 4: Continuous Random Variables"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```


## R Preparation

### Loading packages

Packages or libraries in R provide specialized functions not available in base R.
For example, I have been using the package `learnr` to create these mini-lessons.

Within R, you can call `install.packages("<package name>")` to install most of
the common packages and `library(<package name>)` to load the functions available
in the package.

For example, we have using the following command to install `learnr`.

```{r eval=FALSE, echo=TRUE}
install.packages("learnr")
```

In my code for the lessons, I load the package with the following.

```{r eval=FALSE, echo=TRUE}
library(learnr)
```



### Plotting using `ggplot`

You can create plots with base R without any package.

```{r echo=TRUE}
x <- 1:10
y <- sqrt(x)
plot(x, y)
lines(x, y)
```

Alternatively, `ggplot2` is a popular package for creating plots. I will provide
simple demonstrations of using `ggplot2`. You can learn more about the package
on package [website](https://ggplot2.tidyverse.org).

The code blow

0. `library(ggplot2)`: loads the library.
1. `ggplot(mapping = aes(x = x, y = y))`: creates a `ggplot` canvas and maps 
the vector `x` to `x` axis scale and vector `y` to `y` axis scale in `ggplot`.
2. `theme_classic()`: sets the overall look of the plot to the "classic" theme. I often use `theme_minimal()` as well.
3. `geom_point()`: adds points as specified by the mapping provided in `aes()`. 
4. `geom_line()`: adds a line as specified by the mapping provided in `aes()`.
5. `labs(title = "y is square root of x", x = "x", y = "square root of x")`: customizes the plot and axes titles.
6. `+`: adds the layers together.

```{r echo=TRUE}
library(ggplot2)
ggplot(mapping = aes(x = x, y = y)) +
  theme_classic() +
  geom_point() +
  geom_line() +
  labs(title = "y is square root of x", x = "x", y = "square root of x")
```

Note that the mapping may be provided as you create the plot within `ggplot()`
for all following `geom_*()` calls or override within each `geom_*()`. You
can also change certain aesthetics within each `geom_*()`. 

```{r echo=TRUE}
z <- log(x)
ggplot(mapping = aes(x = x, y = y)) +
  theme_classic() +
  geom_point(color = "blue", size = 3, shape = 1) +
  geom_line() +
  geom_point(aes(y = z), size = 5, colour = "red") +
  geom_line(aes(y = z), linewidth = 1.5, linetype = 2) +
  labs(title = "z is log of x", x = "x", y = "y and z")
```


In the code chunk below, plot `day_of_week` on x-axis vs. `mood` on y-axis with 
a line and points. Change the look of the plot as you see fit.


```{r plot-ex, exercise=TRUE, exercise.lines=14}
day_of_week <- 1:7
mood <- ifelse(day_of_week == 7, 3, day_of_week)


```

```{r plot-ex-solution}
day_of_week <- 1:7
mood <- ifelse(day_of_week == 7, 3, day_of_week)
ggplot(mapping = aes(x = day_of_week, y = mood)) +
  theme_minimal() +
  geom_line() +
  geom_point() +
  labs(x = "", y = "Mood level")

```


## Computing Probabilities for Normal Random Variables

Recall example with $X\sim N\left(1, 4^2\right)$. Use appropriate R functions
to compute the following probabilities.

### $P(X > 2)$


```{r zone, exercise=TRUE}

```

```{r zone-hint-1}
1 - pnorm(1/4) # standardizing using (X - 1) / 4
```
```{r zone-hint-2}
1 - pnorm(2, 1, 4) # without standardizing
```
```{r zone-hint-3}
pnorm(2, 1, 4, lower.tail = FALSE) # directly computing the complement of F
```


### $P(1 < X \le 5)$

```{r ztwo, exercise=TRUE}
```

```{r ztwo-hint-1}
pnorm(1) - pnorm(0) # standardizing using (X - 1) / 4
```
```{r ztwo-hint-2}
pnorm(5, 1, 4) - pnorm(1, 1, 4) # without standardizing
```

### $q_{0.5}$

```{r zq, exercise=TRUE}

```

```{r zq-hint-1}
qnorm(0.5) * 4 + 1 # standardizing using (X - 1) / 4
```
```{r zq-hint-2}
qnorm(0.5, 1, 4) # without standardizing
```


## Computing Probabilities Directly

Recall the local bank drive-up window example with $X\sim \text{Exp}(\lambda_0)$. Suppose $\lambda_0 = 1/5$. Can we compute the pdf and cdf without the internal R distribution functions?

### Compute $f(2)$

$$f(x)=\lambda_0 e^{-\lambda_0 x}$$

```{r expdf, exercise=TRUE, exercise.lines=5}
x <- 2
lambda_0 <- 1/5
```

```{r expdf-solution}
x <- 2
lambda_0 <- 1/5
lambda_0 * exp(-lambda_0 * x)
```



### Compute $F(2)$

$$F(a) = 1 - e^{-\lambda_0 a}$$

```{r expcdf, exercise=TRUE, exercise.lines=5}
a <- 2
lambda_0 <- 1/5
```

```{r expcdf-solution}
a <- 2
lambda_0 <- 1/5
1 - exp(-lambda_0 * a)
```


There may not be R functions for uncommon distributions. Their probabilities can be calculated if the cdf or pmf/pdf are known. 

### Verify

Use `ggplot()` to plot `darkgreen` line for the pdf of $Exp(1/5)$ over values 0 to 10 (`x`)
using `dexp()`. Add `red` points using your version of the function evaluated at integers from 0 to 10 (`w`).


```{r plotpdf, exercise=TRUE, exercise.lines=15}
lambda_0 <- 1/5
x <- seq(0, 10, 0.01)
w <- 0:10

```

```{r plotpdf-hint-1}
lambda_0 <- 1/5
x <- seq(0, 10, 0.01)
w <- 0:10
# both functions output a vector
cdf_R <- pexp(x, lambda_0) 
cdf_diy <- 1 - exp(-lambda_0 * x)
```

```{r plotpdf-solution}
lambda_0 <- 1/5
x <- seq(0, 10, 0.01)
w <- 0:10
# both functions output a vector
cdf_R <- pexp(x, lambda_0) 
cdf_diy <- 1 - exp(-lambda_0 * w)

ggplot() +
  theme_minimal() +
  geom_line(aes(x = x, y = cdf_R), colour = "darkgreen") +
  geom_point(aes(x = w, y = cdf_diy), colour = "red")
```

> `seq(0, 10, 0.01)` creates a vector of values from `0` to `10` increasing by `0.01`. 
Using a small step size makes the line smooth.

Repeat the steps above for the cdf.

```{r plotcdf, exercise=TRUE, exercise.lines=15}
lambda_0 <- 1/5
x <- seq(0, 10, 0.01)
w <- 0:10

```

```{r plotcdf-hint-1}
lambda_0 <- 1/5
x <- seq(0, 10, 0.01)
w <- 0:10
# both functions output a vector
pdf_R <- dexp(x, lambda_0) 
pdf_diy <- lambda_0 * exp(-lambda_0 * x)
```

```{r plotcdf-solution}
lambda_0 <- 1/5
x <- seq(0, 10, 0.01)
w <- 0:10
# both functions output a vector
pdf_R <- dexp(x, lambda_0) 
pdf_diy <- lambda_0 * exp(-lambda_0 * w)

ggplot() +
  theme_minimal() +
  geom_line(aes(x = x, y = pdf_R), colour = "darkgreen") +
  geom_point(aes(x = w, y = pdf_diy), colour = "red")
```
