---
title: "Lecture 4: Continuous Random Variables"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---

<!-- 
Before running the document, make sure `learnr` package is available. 
If missing, you can install the package with `install.packages('learnr')`.
-->

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```


## Computing Probabilities for Normal Random Variables

Recall example with $X\sim N\left(1, 4^2\right)$. Let $Z\sim N(0,1)$.

### Compute $P(X > 2)$

$$P(X>2)=P\left(\frac{X-1}{4} > \frac{2-1}{4}\right)=P\left(Z>\frac{1}{4}\right)$$

```{r zone, exericse=TRUE, exercise.lines=15}

```

### Compute $P(X\le -0.5)$

$$P(X\le -0.5)=P\left(Z\le -\frac{3}{8}\right)$$

```{r ztwo, exercise=TRUE, exercise.lines=15}

```

### Compute $q_{0.25}$

$$q_{0.25}=F^{-1}(0.25)$$

```{r zq, exercise=TRUE, exercise.lines=15}
```


## Computing Probabilities Directly

Recall the local bank drive-up window example with $X\sim \text{Exp}(\lambda_0)$. Suppose $\lambda_0=1$. Can we compute the pdf and cdf without the internal R distribution functions?

### Compute $f(2)$

$$f(x)=\lambda e^{-\lambda x}$$

```{r expd, exercise=TRUE, exercise.lines=20}

```

### Compute $F(2)$

$$F(a) = 1 - e^{-\lambda a}$$

```{r expc, exercise=TRUE, exercise.lines=20}

```


There may not be R functions for uncommon distributions. Their probabilities can be calculated if the cdf or pmf/pdf are known. 



